Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-12-02 paint-contour-tree.tex
done with pass

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to an almost linear (inverse Ackermann overhead) time algorithm for contour tree computation for many
classes of natural instances. 
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
% For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
% and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour. 
% 
It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. More generally, these are the events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.
Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and which form the vertices of the contour tree. 
An edge of the contour tree connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
For example, these could be events where contours gain or lose handles.

Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
We assume the input is $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
$T$ has maximum degree $3$. The degree assumption simplifies the proof and presentation, and is commonly made~\cite{kobps-ctsssit-97}.
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

% \Sesh{Almost everything below is new/reordered. Please read through.}
But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need to entire sorted order of critical points
to construct the contour tree. Is it possible to beat the $t\log t$ bound for these examples?

@@

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
can be computed in $O(\sum_{v \in C(T)} \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of points that are ancestor-descendant in $T$.
\end{theorem}

@@

Note that \Thm{main-corr} is a direct corollary of this statement. This is strictly stronger,
since for any balanced contour tree, the run time bound is $O(t\alpha(t) + N)$, and $O(t)$
comparisons are made.

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value of $t$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
comparisons in the worst-case over this set.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.

There are numerous algorithms for computing contour trees.
An $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, was presented by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and its contour tree $\cC(\MM)$.

Contour trees are a special case of Reeb graphs, a general topological representation for real-valued functions
on any manifold. We refer the reader to Chapter 6 in~\cite{HaEd10} for more details. Algorithms for computing Reeb graphs
is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
two results explicitly reduce to computing contour trees~\cite{TiGySi09,DoNa13}.

@@

We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
and all other values are obtained by linear interpolation.

We assume that $f$ has distinct values on all vertices, except for boundaries,
as expressed in the following constraint, which intuitively requires that 
all boundaries look like (local) horizontal cuts.

@@

We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
where $f(x)$ is the ``height" of a point $x \in \MM$, which is encoded in the representation of $\MM$. 
Specifically, rather than writing our input as $(\MM,f)$, we abuse notation and typically just write $\MM$ to denote the lifted complex.

@@

\begin{definition} \label{def:deg} The \emph{Morse up-degree} of $x$ is the number of $(f(x) + \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The \emph{Morse down-degree} is the number of $(f(x) - \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.
\end{definition}

We categorize points depending on the local changes in topology.

\begin{definition} \label{def:points} A point $x$ is categorized as follows.
\begin{asparaitem}
	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
	\item \emph{Maximum:} Morse up-degree is $0$.
	\item \emph{Minimum:} Morse down-degree is $0$.
	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
\end{asparaitem}
Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
Maxima and minima are \emph{extrema}.
\end{definition}

@@

\newcommand{\newintuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
Analogously, we can define maximum dominant complexes, and both are collectively called extremum dominant.

We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse of the dry part. Each additional cut increases the complexity of the surface,
so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
upwards and downwards.  We finally get a collection of extremum dominant pieces.
We prove a simple contour surgery theorem that builds the contour tree of $\MM$
from the contour trees of the various pieces created.

Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}.
For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
except at a unique boundary face (think of a large boundary triangle that is the boundary).
Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events.

\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
Denote the input as a terrain $\NN$.
The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of higher dimensional faces.} That is, this process is restricted to the 1-skeleton of $\NN$.
Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
paint to reach it.  We begin with an initial painting, since all maximum
colors are distinct.

Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both mounds are colored completely blue or red, while $v$ is touched by both colors.
So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$, and is a critical point.

To process $v$, we ``merge" the colors red and blue into a new color, purple. 
In terms of $\jc(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
Alternatively, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\jc(\MM)$, $\ell_1$ and $\ell_3$
are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.

@@

What happens when finally blue and red join at $v$? We merge the two colors, but now have
blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
This necessitates using a priority queue with efficient merges.
Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
There are numerous details to handle to prove the final run time bound, but this should provide the reader
some intuition behind our algorithm.
}

@@

The following claim gives a description of the interface. While its meaning is quite intuitive, the proof is tedious.

@@

We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.

@@

\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
%
%
%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
%contradicting the fact that $\phi$ is dry.
\end{proof}

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complexes. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
of the procedure is that the rain alternately flows downwards and upwards, 
since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
In other words, when
we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
opposite direction. While one can implement $\rain$ this way, it is conceptually easier
to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
is easily achieved by just negating the height values. (From a running time standpoint, it suffices
to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

@@

\medskip
For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
We first argue the correctness of $\rain$.

@@

@@

\Ben{Read remainder of this section carefully, it has been cut up, and parts rewritten to fit.}

@@

A painting colors edges, not vertices. We associate certain
sets of colors with vertices.

>>>>

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
basically $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
This matches al existing upper bounds, but is a significant improvement when the contour tree is short and fat.
% Our results match all existing upper bounds, but also
% characterize the amount of sorting required to compute the contour tree.
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
% For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
% and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour. 
% 
It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. 
The contour tree is a concise representation of all these events.
Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
An edge of the contour tree connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.) 
\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
and internal nodes are either ``joins" or ``splits".

% so we can talk of the ``up-degree" and ``down-degree" of vertices.
% Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
% $1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
% merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
% In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
% For example, these could be events where contours gain or lose handles.

Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
$T$ has maximum degree $3$. The degree assumption simplifies the presentation, and is commonly made~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

% \Sesh{Almost everything below is new/reordered. Please read through.}
All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need to entire sorted order of critical points
to construct the contour tree.

@@

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
can be computed in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of points that are ancestor-descendant in $T$.
\end{theorem}

@@

Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at most
the length of the path in $P(T)$ that contains $v$.
This bound is strictly stronger,
since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(t)$
comparisons are made.

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value of $t$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
comparisons in the worst-case over this set. (All further details are 
in Appendix~\Sec{lb}.)

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
Contour trees have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
% Carr's thesis shows relationships between monotone paths
% in $\MM$ and its contour tree $\cC(\MM)$.

Contour trees are a special case of Reeb graphs, a general topological representation for real-valued functions
on any manifold. Algorithms for computing Reeb graphs
is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
two results explicitly reduce to computing contour trees~\cite{TiGySi09,DoNa13}.

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is explicitly defined only on the vertices of $\MM$,
and all other values are obtained by linear interpolation.

We assume that the boundary values satisfy a special property. This is mainly for convenience
in presentation.

@@

We think of the dimension $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
where $f(x)$ is the ``height" of a point $x \in \MM$, which is encoded in the representation of $\MM$. 
Specifically, rather than writing our input as $(\MM,f)$, we abuse notation and typically just write $\MM$ to denote the lifted complex.

@@

\begin{definition} \label{def:deg} The \emph{Morse up-degree} of $x$ is the number of $(f(x) + \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The \emph{Morse down-degree} is the number of $(f(x) - \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.

A \emph{regular} point has both Morse up-degree and down-degree $1$. A \emph{maximum} has Morse up-degree $1$, while
a \emph{minimum} has Morse up-degree $2$. A \emph{Morse Join} has Morse up-degree strictly greater than $1$,
while a \emph{Morse Split} has Morse down-degree strictly greater than $1$. Non-regular points are called \emph{critical}.
\end{definition}

% We categorize points depending on the local changes in topology.
% 
% \begin{definition} \label{def:points} A point $x$ is categorized as follows.
% \begin{asparaitem}
% 	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
% 	\item \emph{Maximum:} Morse up-degree is $0$.
% 	\item \emph{Minimum:} Morse down-degree is $0$.
% 	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
% 	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
% \end{asparaitem}
% Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
% Maxima and minima are \emph{extrema}.
% \end{definition}
%

@@

\newcommand{\newintuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a short, high-level
description of the main ideas.
% However, for the $d=2$ case, 
% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
% can be explained clearly.

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse of the dry part. This is carefully done to ensure that the total complexity
of the pieces is linear.
% 
% 
% Each additional cut increases the complexity of the surface,
% so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
% upwards and downwards.  We finally get a collection of extremum dominant pieces.
We prove a simple contour surgery theorem that builds the contour tree of $\MM$
from the contour trees of the various pieces created.

Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}.
% For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
% except at a unique boundary face (think of a large boundary triangle that is the boundary).
Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events.

\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
paint to reach it.

Our algorithm incrementally builds the join tree from the leaves (maxima) to the root. 
Refer to the left part of \Fig{colors}.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both mounds are colored completely blue or red, while $v$ is touched by both colors.
So this indicates that $v$ joins the blue maximum and red maximum in the join tree and is a critical point.
To process $v$, we ``merge" the colors red and blue into a new color, purple. 
In terms of join tree, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. 
A critical insight is the use of \emph{binomial heaps}~\cite{Vu78} to handle all the priorities and the merging.
This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree.
There are numerous details to handle to prove the final run time bound, but this should provide the reader
some intuition behind our algorithm.

@@

% What happens when finally blue and red join at $v$? We merge the two colors, but now have
% blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
% This necessitates using a priority queue with efficient merges.
% Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
% We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
}

@@

The following claim gives a description of the interface.

@@

We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
The following claim follows directly from \Thm{jordan}.

@@

\InNotSoCGVer{
\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
%
%
%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
%contradicting the fact that $\phi$ is dry.
\end{proof}
}

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complexes. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
% One of the critical aspects
% of the procedure is that the rain alternately flows downwards and upwards, 
% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
When
we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
opposite direction. This is crucial to ensure a linear running time. 
The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
% While one can implement $\rain$ this way, it is conceptually easier
% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

@@

\medskip
For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
% We first argue the correctness of $\rain$.

@@

The following are direct consequences of \Lem{new-verts} and the surgery procedure.

@@

@@

% A painting colors edges, not vertices. We associate certain
% sets of colors with vertices.
